Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-08-27 analysis/paper/paper.Rmd
fixing lamb ref, used visual mode, RStudio changed formatting

The number of species threatened with extinction is high and getting higher [@ipbes_global_2019]. Reliable information on the status of populations and likely impacts of management are needed to inform conservation action but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018;@wintle_allocating_2010;@dunham_strategic_2023]. Resources to support conservation and recovery are limited [@buxton_balancing_2022;@mccarthy_financial_2012;@gerber_conservation_2016;@wintle_spending_2019], delays increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013]. A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. Analysis of the implications of monitoring outcomes for decisions can help clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].

Analysis methods that can help make better use of available monitoring data to inform decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012]. To be useful, these methods must be tailored to particular species and decision contexts. Here, we offer a simple Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].  Following the value of information analysis framework proposed by Dunham et al. (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020]. The causes of these demographic-disturbance relationships include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011]. Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].

@@

The most widely available source of boreal caribou demographic information is from cows fitted with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@decesare_estimating_2012], though genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013, @eacker_web-based_2019, and @lamb_effectiveness_2024 for examples of application to caribou). Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa , and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020]. Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effective_2024], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018]. Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019;@lamb_effectiveness_2024] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.

We propose to reduce uncertainty in assessments and projections of local caribou population viability by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data in a simple Bayesian population model for boreal caribou [building on @eacker_web-based_2019]. In our simple proof-of-concept model, information is shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. A more rigorous statistical approach would be to integrate multiple populations into a hierarchical Bayesian analysis of variation within and among populations [@leasure_hierarchical_2019]; we have not done this because owners have not authorized the use of the data for this purpose, obtaining permission to use caribou demographic data is difficult, and most managers only have access to their own local data. Despite limitations, our simple Bayesian model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships. We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes. We hope that this approach can help clarify where available information is sufficient for assessing and projecting population viability, and thereby reduce the risk that uncertainty is used as a reason to delay conservation action for imperiled populations.

While we believe that using available information to reduce uncertainty is important, we also believe it is important to acknowledge remaining uncertainty. In general, parametric Bayesian population models include a number of simplifying assumptions, and misspecified models can yield biased or overconfident results [@schaub_integrated_2021]. More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021]. Although several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012] none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019]. We address this gap with a bias correction term derived from a simple model of the recruitment survey observation process.

@@

We focus on exploring a set of scenarios in which
buffered anthropogenic disturbance, measured as the percentage of area within 500
m of some type of anthropogenic disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through medium-low (20\%) and medium-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

\begin{equation}
\bar{R}_t \sim Beta(\mu^R_t,\phi^R); 
log(\mu^R_t) = \dot{\beta}^R_0+\dot{\beta}^R_a A_t+\dot{\beta}^R_f F_t,
  (\#eq:barR)
\end{equation}

\begin{equation}
\bar{S}_t \sim (46 \times Beta(\mu^S_t,\phi^S)-0.5)/45;
log(\mu^S_t) = \dot{\beta}^S_0+\dot{\beta}^S_a A_t.
  (\#eq:barS)
\end{equation}

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations
(Figure \@ref(fig:exampleTrajectories)). To ensure results are comparable across monitoring scenarios and that populations remain extant through the monitoring period, all simulations are started with $N_0=5000$ animals 24 years prior to the start of the projection period.

@@

Following others [@eacker_web-based_2019;@lamb_effectiveness_2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a Gaussian
distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019,@lamb_effectiveness_2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

<!-- Annotated equations for poster.
$$J_{t} \sim \text{Poisson}(W_tX_t)$$
$$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
$$W_{t} \sim \text{Binomial}(N_t,\underbrace{S_t}_{\text{Adult survival probability}})$$
$$S_t=\underbrace{\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})}_{\text{Adjustment as in in Johnson et al. 2020}}=\beta^S_0+ \underbrace{\beta^S_aA_t}_{\text{Anthropogenic disturbance effect}}+\underbrace{\epsilon^S_t}_{\text{Random effect of year}}.$$
$$J_{t} \sim \text{Poisson}(W_t\underbrace{R_t/2}_{\text{Recruitment rate}})$$
$$\log(R_t)=\beta^R_0+\underbrace{\beta^R_a A_t}_{\text{Anthropogenic disturbance effect}}+ \underbrace{\beta^R_fF_t}_{\text{Fire effect}}+\underbrace{\epsilon^R_t}_{\text{Random effect of year}}.$$
$$\underbrace{\hat{C}_t}_{\text{Observed calves}} \sim \text{Binomial}(\underbrace{\hat{W}_t}_{\text{Observed cows}},R_t)$$
$$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk. To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
\*
see alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
-->

Models and analysis were implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse_ packages [@wickham_welcome_2019]. We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model. We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2. We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018]. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>, and in the  _caribouMetrics_ R package available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of perfect information (EVPI) and the expected value of sample information (EVSI) [@runge_which_2011;@yokota_value_2004]. We assign a value or utility of $\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$ matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$ when the status assessment is incorrect. This is "the value ascribed by the decision-maker to the outcome, and is thus the measure of management performance" [@runge_which_2011]. The expected value of current information is the average utility of decisions informed by knowledge of the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model:
\begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. The expected value under certainty (i.e. the expected value of a decision when the true state of the population is known) is $EV_{certainty}=1$. The difference between $EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum possible benefit that could theoretically be gained if the system were perfectly known.

@@

The availability of boreal caribou demographic data varies widely across Canada due to large variation in habitat condition, population condition, available resources for monitoring, and remoteness. In a few regions enough data has been collected over time to characterize local demographic-disturbance relationships [@rudolph_demographic_2017;@hervieux_widespread_2013], but most populations have only been sparsely monitored for snapshot estimates of trend or abundance [@johnson_science_2020]. As a result, in most regions it is necessary to make assumptions informed by observed variation in demographic rates with disturbance among populations and other relevant information about impacts of disturbance on caribou. However, high observed variation in demography where anthropogenic disturbance is low leads to high uncertainty in projections informed by national demographic-disturbance relationships alone [@stewart_climate-informed_2023;@dyson_existing_2022]. In some cases, predictions of population growth rate informed by national demographic-disturbance relationships appear to disagree with estimates from local demographic data [@eccc_scientific_2011; @dyson_existing_2022], but without formal methods of accounting for uncertainty it can be difficult to distinguish true discrepancies from stochastic variation.

We developed a simple Bayesian population model [building on @eacker_web-based_2019] to help address these challenges by acknowledging uncertainty associated with local demographic data and national demographic-disturbance relationships, and integrating these two sources of information to reduce uncertainty in projections of the impacts of changing landscape conditions. This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.

>>>>

The number of species threatened with extinction is high and getting
higher [@ipbes_global_2019]. Reliable information on the status of
populations and likely impacts of management are needed to inform
conservation action but effective monitoring is challenging and costly
[@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
@dunham_strategic_2023]. Resources to support conservation and recovery
are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
@gerber_conservation_2016; @wintle_spending_2019], delays increase risk
[@martin_acting_2012], and some populations have been monitored to
extinction [@lindenmayer_counting_2013]. A recent analysis of action
plans for Canadian species at risk found half of tasks are research and
monitoring, and this proportion is highest for species with higher risk
of extinction, highlighting a need to more carefully consider when,
where, and why additional information is needed to inform conservation
action [@buxton_balancing_2022]. Analysis of the implications of
monitoring outcomes for decisions can help clarify whether additional
monitoring is necessary to inform conservation action
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012].

Analysis methods that can help make better use of available monitoring
data to inform decisions include Bayesian population modelling
[@schaub_integrated_2021] and value of information analyses
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012]. To be useful, these methods must be tailored
to particular species and decision contexts. Here, we offer a simple
Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates information about the impacts of landscape
change with available demographic data to reduce uncertainty in
population viability projections for boreal caribou [building on
@eacker_web-based_2019]. Following the value of information analysis
framework proposed by Dunham et al. (2023), we combine this Bayesian
model with simulations of plausible population trajectories
[@stewart_climate-informed_2023; @dyson_existing_2022] to show how the
need for local monitoring varies with landscape condition, and to assess
the ability of alternative monitoring scenarios to reduce the risk of
errors in population viability projections. Our approach is tailored to
boreal caribou, but could be adapted for other species at risk found in
multiple populations that are costly to monitor and sensitive to habitat
disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian
boreal region across a gradient of anthropogenic disturbance
[@eccc_scientific_2011; @johnson_science_2020] that includes some of the
most intact and inaccessible places remaining on earth
[@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both
survival and recruitment decrease with the percentage of the study area
that is disturbed, and no populations of caribou are considered
self-sustaining in study areas with high anthropogenic disturbance
[@eccc_scientific_2011; @johnson_science_2020]. The causes of these
demographic-disturbance relationships include direct impacts on habitat
availability and use, and indirect impacts on predators and alternate
prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreases in
recruitment as disturbance increases are also observed over time in
places where there is sufficient historic demographic data
[@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
lack sufficient data to characterize local demographic-disturbance
relationships [@johnson_science_2020]. There is also variation in
demographic rates among study areas, particularly where anthropogenic
disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], and this variation leads to
uncertainty in projections derived from national disturbance-demographic
relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

@@

The most widely available source of boreal caribou demographic
information is from cows fitted with telemetry collars [e.g.
@courtois_aerial_2003; @rettie_dynamics_1998; @hervieux_widespread_2013;
@decesare_estimating_2012], though genetic methods are playing an
increasingly important role in some caribou monitoring programs [e.g.
@hettinga_estimating_2012; @mcfarlane_genetic_2018;
@moeller_integrating_2021; @flasko_fecal_2017; @jones_novel_2023;
@mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows
are typically captured via net-gun from aircraft in late winter, and
affixed with with collars that send a mortality signal when no longer
active [see @nbckc_nbckc_2021], thus allowing estimation of adult
survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for
original approach, and @mcloughlin_declines_2003,
@hervieux_widespread_2013, @eacker_web-based_2019, and @lamb2024 for
examples of application to caribou). Recruitment is estimated from
aerial surveys of the composition of groups of animals that include
collared animals [@nbckc_nbckc_2022]. Groups typically also contain
uncollared animals, and a critical assumption in these surveys is that
different demographic groups are equally detectable
[@ellington_divergent_2020]. Errors that can bias composition survey
results include misidentifying young bulls as cows or vice versa , and
failing to detect calves [@ellington_divergent_2020;
@decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple
recruitment-mortality population model [@hatter_moose_1991] that serves
as the basis for many herd viability assessments
[@decesare_estimating_2012; @serrouya_comparing_2017;
@eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].
Some efforts have been made to improve caribou population viability
projections and growth rate estimates using Bayesian methods
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb2024], following broader trends in the
analysis of population dynamics [@schaub_integrated_2021]. In Bayesian
population models, all available demographic or population data (e.g.
abundance surveys, vital rate estimates, trend data) are combined into a
single analysis that can permit novel parameter estimation, and improve
precision, and account for multiple sources of uncertainty
[@schaub_integrated_2011; @eacker_assessing_2017;
@schaub_integrated_2021]. These models are proving broadly useful for
rapid detection of population declines, evaluation of management
actions, and viability analysis of critically endangered species
[@schaub_integrated_2021]. However, there is very little monitoring data
available for some caribou populations [@johnson_science_2020] and there
is a risk that lack of local data will be used as a reason to delay
conservation action [@boan_climate_2018]. Bayesian caribou population
models [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @lamb2024] have not thus far included
demographic-disturbance relationships, and we are concerned that models
informed only by local demographic data could be used to inflate
uncertainty about the status of poorly monitored populations.

We propose to reduce uncertainty in assessments and projections of local
caribou population viability by integrating prior information from
national analysis of demographic-disturbance relationships with
available local demographic data in a simple Bayesian population model
for boreal caribou [building on @eacker_web-based_2019]. In our simple
proof-of-concept model, information is shared among populations using
priors derived from a published analysis of variation in demographic
rates among populations across Canada [@johnson_science_2020]. A more
rigorous statistical approach would be to integrate multiple populations
into a hierarchical Bayesian analysis of variation within and among
populations [@leasure_hierarchical_2019]; we have not done this because
owners have not authorized the use of the data for this purpose,
obtaining permission to use caribou demographic data is difficult, and
most managers only have access to their own local data. Despite
limitations, our simple Bayesian model represents an important
improvement over less formally integrated methods for estimating and
projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

While we believe that using available information to reduce uncertainty
is important, we also believe it is important to acknowledge remaining
uncertainty. In general, parametric Bayesian population models include a
number of simplifying assumptions, and misspecified models can yield
biased or overconfident results [@schaub_integrated_2021]. More
specifically, models that do not account for observation errors can be
misleading [@schaub_integrated_2021]. Although several authors have
pointed out that misclassification of adults and missing calves in
composition surveys can bias recruitment estimates from those surveys
[@ellington_divergent_2020; @decesare_estimating_2012] none have thus
far examined sensitivity of Bayesian caribou population models to these
sources of error [@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019]. We address this gap with a bias correction term
derived from a simple model of the recruitment survey observation
process.

@@

We focus on exploring a set of scenarios in which buffered anthropogenic
disturbance, measured as the percentage of area within 500 m of some
type of anthropogenic disturbance [@johnson_science_2020;
@eccc_scientific_2011], increases by 1% per year (Figure
\@ref(fig:figDisturbance)). The duration of monitoring prior to the
projection period varies among scenarios from 1 to 24 years. The
projection period begins in 2024, and we consider how uncertainty
changes as the projection period increases from 5 to 20 years. The
amount of anthropogenic disturbance at the beginning of the projection
period varies from low (0%) through medium-low (20%) and medium-high
(40%) to high (60%). These simple hypothetical scenarios span a large
range to show how the need for local monitoring to reduce uncertainty
depends on landscape state, not to represent the reality of landscape
change. Scenarios in which anthropogenic disturbance remains the same or
decreases over time are included in Supplement D.

@@

```{=tex}
\begin{equation}
\bar{R}_t \sim Beta(\mu^R_t,\phi^R); 
log(\mu^R_t) = \dot{\beta}^R_0+\dot{\beta}^R_a A_t+\dot{\beta}^R_f F_t,
  (\#eq:barR)
\end{equation}
```
```{=tex}
\begin{equation}
\bar{S}_t \sim (46 \times Beta(\mu^S_t,\phi^S)-0.5)/45;
log(\mu^S_t) = \dot{\beta}^S_0+\dot{\beta}^S_a A_t.
  (\#eq:barS)
\end{equation}
```
$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period.

@@

Following others [@eacker_web-based_2019; @lamb2024], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the *survival* R package [@therneau_modeling_2000],
with precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is modelled as a
Gaussian distributed function of the true survival probability,
truncated between 0 and 1, to allow sharing survival information across
years with a random effect [@eacker_web-based_2019,@lamb2024]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

@@

```{=html}
<!-- Annotated equations for poster.
$$J_{t} \sim \text{Poisson}(W_tX_t)$$
$$\underbrace{N_{t+1}}_{\text{Adult females in year t+1}}=\underbrace{W_t}_{\text{Surviving adults}}+\underbrace{J_t}_{\text{New recruits to adult age class}}$$
$$W_{t} \sim \text{Binomial}(N_t,\underbrace{S_t}_{\text{Adult survival probability}})$$
$$S_t=\underbrace{\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})}_{\text{Adjustment as in in Johnson et al. 2020}}=\beta^S_0+ \underbrace{\beta^S_aA_t}_{\text{Anthropogenic disturbance effect}}+\underbrace{\epsilon^S_t}_{\text{Random effect of year}}.$$
$$J_{t} \sim \text{Poisson}(W_t\underbrace{R_t/2}_{\text{Recruitment rate}})$$
$$\log(R_t)=\beta^R_0+\underbrace{\beta^R_a A_t}_{\text{Anthropogenic disturbance effect}}+ \underbrace{\beta^R_fF_t}_{\text{Fire effect}}+\underbrace{\epsilon^R_t}_{\text{Random effect of year}}.$$
$$\underbrace{\hat{C}_t}_{\text{Observed calves}} \sim \text{Binomial}(\underbrace{\hat{W}_t}_{\text{Observed cows}},R_t)$$
$$\underbrace{\hat{S_t}}_{\text{Observed survival (Kaplan-Meier estimate)}} \sim \text{Normal}(S_t,1/\tau_t).$$ -->
```
```{=html}
<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives biased results if there is left censoring, or months without animals at risk. To address this we drop left censored observations and only estimate survival when at least some animals remain at risk by month 12.
\*
see alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. See also Lamb et al, who uses two step KM estimation process for survival like Eacker. Stick with published examples. \*
-->
```
Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

As suggested by Dunham et al. [-@dunham_strategic_2023] we also
calculate the expected value of perfect information (EVPI) and the
expected value of sample information (EVSI) [@runge_which_2011;
@yokota_value_2004]. We assign a value or utility of
$\nu_{\alpha,\psi}=1$ when the true state of the population $\psi$
matches the assigned status $\alpha$, and value of $\nu_{\alpha,\psi}=0$
when the status assessment is incorrect. This is "the value ascribed by
the decision-maker to the outcome, and is thus the measure of management
performance" [@runge_which_2011]. The expected value of current
information is the average utility of decisions informed by knowledge of
the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and
priors from the national model: \begin{equation}
EV_{current} = max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVcurrent)
\end{equation} where $P(\psi|\tau)$ is the prior probability that the
true state of the population is $\psi$, and each population is assigned
a status $\alpha$ informed by the prior. The expected value under
certainty (i.e. the expected value of a decision when the true state of
the population is known) is $EV_{certainty}=1$. The difference between
$EV_{certainty}$ and $EV_{current}$ gives EVPI, a measure of the maximum
possible benefit that could theoretically be gained if the system were
perfectly known.

@@

The availability of boreal caribou demographic data varies widely across
Canada due to large variation in habitat condition, population
condition, available resources for monitoring, and remoteness. In a few
regions enough data has been collected over time to characterize local
demographic-disturbance relationships [@rudolph_demographic_2017;
@hervieux_widespread_2013], but most populations have only been sparsely
monitored for snapshot estimates of trend or abundance
[@johnson_science_2020]. As a result, in most regions it is necessary to
make assumptions informed by observed variation in demographic rates
with disturbance among populations and other relevant information about
impacts of disturbance on caribou. However, high observed variation in
demography where anthropogenic disturbance is low leads to high
uncertainty in projections informed by national demographic-disturbance
relationships alone [@stewart_climate-informed_2023;
@dyson_existing_2022]. In some cases, predictions of population growth
rate informed by national demographic-disturbance relationships appear
to disagree with estimates from local demographic data
[@eccc_scientific_2011; @dyson_existing_2022], but without formal
methods of accounting for uncertainty it can be difficult to distinguish
true discrepancies from stochastic variation.

We developed a simple Bayesian population model [building on
@eacker_web-based_2019] to help address these challenges by
acknowledging uncertainty associated with local demographic data and
national demographic-disturbance relationships, and integrating these
two sources of information to reduce uncertainty in projections of the
impacts of changing landscape conditions. This Bayesian approach
represents an important improvement over less formally integrated
methods for estimating and projecting impacts of disturbance on
population population viability [@eccc_scientific_2011;
@johnson_science_2020] because it reconciles and accounts for
uncertainties in local demographic data and demographic-disturbance
relationships.

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus informed by a number of
lines of evidence that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.
